The Legacy of the European Post-Master in Urbanism at TU Delft: A Text Mining Approach

### 2023-03-13 analysis/paper/paper.qmd
Add notes to the introduction

# Load packages ----
## Managing paths
library(here)

## Data manipulation and visualisation
library(tidyverse)

## Text processing and analysis
library(tidytext)
library(SnowballC)
library(tm)

## Working with spatial data
library(tidygeocoder)

@@

```{r}
#| label: extract-pdf-text
#| echo: false
emu_theses$text <- convert_pdf_text(pdf_paths)
```

@@

emu_theses$text_clean <- emu_theses$text |>
  clean_basic() 
```

@@

In order to describe the legacy of the EMU program, including the distinctive features of its didactic approach, this paper aims to reveal the main topics taught in it and how those topics had evolved through the years of the program. To that end, we employed a text mining approach in which we analysed its output: `r nrow(theses_all)` theses with an average of `r mean(str_length(emu_theses$text_clean), na.rm = TRUE)` words produced over the years for the duration of the program between `r paste0(min(theses_all$grad_year, na.rm = TRUE), "-", max(theses_all$grad_year, na.rm = TRUE))`. The first year of the program was not represented, as PDF files were only available from 2008 onward.

@@

The theses are available in PDF format with complex layouts typical to an urbanism project in which text and various types of graphics are combined.

A total number of `r nrow(emu_theses)` were analysed.

@@

# Show thesis locations on world map
world <- map_data("world")
ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.1
  ) +
  geom_point(data = emu_theses, aes(longitude, latitude), color = "red") +
  coord_fixed() +
  theme_void()
```

@@

emu_theses$text_clean <- emu_theses$text_clean |>
  str_remove_all(find_meta_stopwords(emu_theses)) |> 
  str_remove_all(subject_stopwords()) |> 
  str_remove_all(document_stopwords())

emu_theses_words <- emu_theses |> 
  select(title, text_clean) |> 
  unnest_tokens(output = word, input = text_clean) |>   # remove punctuation, convert to lowercase, separate all words
  anti_join(tidytext::stop_words, by = "word")
```

@@

```{r}
# Which are the top 20 most frequently used words in each thesis?
emu_theses_top_20 <- emu_theses_words %>%
  group_by(title) %>%
  count(word, sort = TRUE) %>%
  slice_max(n, n = 20) %>%        ### top_n is superseded; better use slice_max instead
  # filter(n > 50) %>%              ### filter words with a minimum count
  ungroup() %>% 
  mutate(word = reorder(word, n))

# Which are first 5 most frequently used words in each thesis?
emu_theses_first_5 <- emu_theses_words %>%
  count(title, word) %>% 
  group_by(title) %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

# Which theses contain "fabric" in the top 20 most frequently used words?
emu_theses_top_20 %>% 
  filter(word == "fabric")

emu_theses_top_20 %>%
  slice(1:40) %>% 
  ggplot(aes(x = reorder(word, n), y = n, fill = title)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
  labs(y = "Count",
       x = "Unique words",
       title = "Count of unique words found in thesis",
       subtitle = "Stop words were removed from the list") +
  facet_wrap( ~ title, scales = "free", ncol = 2)

# # wip function ngrams
# get_ngrams <- function(df, n) {
#   if (!is.numeric(n)) {
#     stop("n is not numeric")
#   }
#   emu_theses %>%
#     mutate(author = paste(first_name, last_name)) %>% 
#     select(author, title, text, -c(year, first_name, last_name, text_start, text_end)) %>% 
#     unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
#     separate(bigram, into = c("first","second"), sep = " ", remove = FALSE) %>%
#     anti_join(stop_words, by = c("first" = "word")) %>%
#     anti_join(stop_words, by = c("second" = "word")) %>%
#     anti_join(stop_words_custom, by = c("first" = "word")) %>%
#     anti_join(stop_words_custom, by = c("second" = "word")) %>%
#     filter(str_detect(first, "[a-z]") & str_detect(second, "[a-z]")) %>% 
#     filter(first != second) %>% 
#     filter(first != "source")
# }
# emu_theses_ngrams <- get_ngrams(emu_theses, 2)

# bigrams
emu_theses_bigrams <- emu_theses %>%
  mutate(author = paste(first_name, last_name)) %>% 
  select(author, title, text, -c(graduation_year, first_name, last_name)) %>% 
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  separate(bigram, into = c("first","second"), sep = " ", remove = FALSE) %>%
  anti_join(stop_words, by = c("first" = "word")) %>%
  anti_join(stop_words, by = c("second" = "word")) %>%
  anti_join(stop_words_custom, by = c("first" = "word")) %>%
  anti_join(stop_words_custom, by = c("second" = "word")) %>%
  filter(str_detect(first, "[a-z]") & str_detect(second, "[a-z]")) %>% 
  filter(first != second) %>% 
  filter(first != "source")

emu_theses_bigrams_top <- emu_theses_bigrams %>%
  group_by(title) %>%
  count(bigram, sort = TRUE) %>%
  slice_max(n, n = 10) %>%        ### top_n is superseded; better use slice_max instead
  # filter(n > 50)                ### filter words with a minimum count
  ungroup() %>% 
  mutate(bigram = reorder(bigram, n))

@@

emu_theses_bigrams_top %>%
  ggplot(aes(x = bigram, y = n, fill = title)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
  labs(y = "Count",
       x = "Unique words",
       title = "Count of unique words found in thesis",
       subtitle = "Stop words were removed from the list") +
  facet_wrap( ~ title, scales = "free", ncol = 2)

# find the most frequently used bigrams
emu_thesis_bigrams_top <- emu_theses_bigrams %>%
  group_by(title) %>% 
  count(bigram, sort = TRUE) %>%
  slice_max(n, n = 10) %>%                   ### change the number of top words to show
  mutate(bigram = reorder(bigram, n)) %>% 
  print()
```

>>>>

# Load packages ----
## List packages to be loaded
packages <- c(
  "here",                                        # Managing paths
  "tidyverse",                                   # Data manipulation and visualisation
  "tidytext", "SnowballC", "tm", "topicmodels",  # Text processing and analysis
  "tidygeocoder",                                # Working with spatial data
  "igraph", "ggraph"                             # Network analysis and visualsation
)

## Load required packages, install them if needed
for (package in packages) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package,  "http://cran.us.r-project.org")
  }
  library(package, character.only = TRUE)
}

# library(here)
# library(tidyverse)
# library(tidytext)
# library(SnowballC)
# library(tm)
# library(topicmodels)
# library(tidygeocoder)
# library(igraph)
# library(ggraph)

@@

```{r}
#| label: extract-pdf-text
#| echo: false

# Extract text from PDFs
# On the first run, this function might take extra time as a Python environment needs to be set up

emu_theses$text <- convert_pdf_text(pdf_paths)
```

@@

emu_theses$text_clean <- emu_theses$text |> 
  clean_basic() |> 
  str_remove_all(find_meta_stopwords(emu_theses) )  
```

```{r}
#| label: preview
#| include: false

# Preview theses in increasing order of graduation year
emu_theses %>% 
  arrange(graduation_year) %>% 
  pull(text_clean) %>% 
  head_text(head_n = 5)

# Show distribution of the length of theses
hist(str_length(emu_theses$text_clean))
```

@@

Distinctive features of the program:  
- as a postmaster, it only accepted applicants with a prior MSc degree and experience in practice
- provides a bridge towards PhD-level research
- all students had an exchange semester at one of the other three program universities

In order to describe the legacy of the EMU program, including the distinctive features of its didactic approach, this paper aims to reveal the main topics taught in it and how those topics had evolved through the years of the program. To that end, we employed a text mining approach in which we analysed its output: `r nrow(all_theses)` theses with an average of `r mean(str_length(emu_theses$text_clean), na.rm = TRUE)` words produced over the years for the duration of the program between `r paste0(min(all_theses$grad_year, na.rm = TRUE), "-", max(all_theses$grad_year, na.rm = TRUE))`. The first year of the program was not represented, as PDF files were only available from 2008 onward.

@@

The theses are available in PDF format with complex layouts typical to an urbanism project in which text and various types of graphics are combined. With the exception of a bilingual English-Spanish thesis, all thesis were written in English.

A total number of `r nrow(emu_theses)` out of `r nrow(all_theses)` were analysed.

@@

# Show thesis locations on world map
world <- ggplot2::map_data("world")
ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.1
  ) +
  geom_count(data = emu_theses, aes(longitude, latitude), color = "red", alpha = 0.7) +
  scale_size_area(max_size = 3) +
  coord_fixed() +
  theme_void()
```

@@

stop_words_custom <- emu_theses %>% 
  unnest_tokens(output = word, input = title) %>%   ### add words from the title
  select(word) %>% 
  rbind(., data.frame(word = c("preface", "foreword", "introduction", "conclusion", "thesis", 
                               "source", "author"))) %>% 
  # rbind(., data.frame(word = c("city", "urban", "urbanism"))) %>% 
  rbind(., data.frame(word = c("hab", "km")))

emu_theses_words <- emu_theses %>%
  mutate(author = paste(first_name, last_name)) %>% 
  select(author, title, text, -c(graduation_year, first_name, last_name)) %>% 
  unnest_tokens(output = word, input = text) %>%   # remove punctuation, convert to lowercase, seperate all words
  anti_join(stop_words, by = "word") %>%  # remove stop words
  anti_join(stop_words_custom, by = "word")
```

@@

```{r}
# Which are the top n most frequently used words in each thesis?
get_top_words_per_document(data = emu_theses_words, 
                           top_n = 3, 
                           title_col = "title", 
                           word_col = "word")
```

```{r}
# Which are the top n most frequently used words in all theses`?
get_top_words_per_corpus(data = emu_theses_words, 
                         top_n = 5, 
                         word_col = "word")
```

```{r}
# Which are the top n most frequently used words in all theses`?
emu_theses_top_20 <- get_top_words_per_corpus(data = emu_theses_words, 
                         top_n = 20, 
                         word_col = "word")
```

<!-- ```{r} -->

<!-- # Which theses contain "fabric" in the top 20 most frequently used words? -->

<!-- emu_theses_top_20 %>%  -->

<!--   filter(word == "fabric") -->

<!-- emu_theses_top_20 %>% -->

<!--   slice(1:40) %>%  -->

<!--   ggplot(aes(x = reorder(word, n), y = n, fill = title)) + -->

<!--   geom_col() + -->

<!--   xlab(NULL) + -->

<!--   coord_flip() + -->

<!--   labs(y = "Count", -->

<!--        x = "Unique words", -->

<!--        title = "Count of unique words found in thesis", -->

<!--        subtitle = "Stop words were removed from the list") + -->

<!--   facet_wrap( ~ title, scales = "free", ncol = 2) -->

<!-- ``` -->

```{r}
#| label: ngrams

## TO DO: generalise bigrams and trigrams into ngrams function

## Bigrams
emu_theses_bigrams <- emu_theses |> 
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |> 
  separate(bigram, into = c("first", "second"), sep = " ", remove = FALSE) |> 
  anti_join(stop_words, by = c("first" = "word")) |> 
  anti_join(stop_words, by = c("second" = "word")) |> 
  filter(str_detect(first, "[a-z]") & 
         str_detect(second, "[a-z]"))

bigram_counts <- emu_theses_bigrams |> 
  count(first, second, sort = TRUE)

bigram_counts

## Trigrams
emu_theses_trigrams <- emu_theses |> 
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |> 
  separate(trigram, into = c("first", "second", "third"), sep = " ", remove = FALSE) |> 
  filter(!first %in% stop_words$word,
         !second %in% stop_words$word,
         !third %in% stop_words$word) |>  
  filter(str_detect(first, "[a-z]"), 
         str_detect(second, "[a-z]"),
         str_detect(third, "[a-z]"))

trigram_counts <- emu_theses_trigrams |> 
  count(first, second, sort = TRUE)

trigram_counts

## Bigram tf-idf
bigram_tf_idf <- emu_theses_bigrams |> 
  count(title, bigram) |> 
  bind_tf_idf(bigram, title, n) |> 
  arrange(desc(tf_idf))

bigram_tf_idf

## Trigram tf-idf
trigram_tf_idf <- emu_theses_trigrams |> 
  count(title, trigram) |> 
  bind_tf_idf(trigram, title, n) |> 
  arrange(desc(tf_idf))

trigram_tf_idf

## Show network of bigrams
bigram_graph <- bigram_counts |> 
  filter(n > 100) |> 
  graph_from_data_frame()

bigram_graph

set.seed(2023)

a <- grid::arrow(type = "closed", length = unit(.15, "inches"))

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), show.legend = FALSE, arrow = a) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
  theme_void()

## Higgh frequency bigrams ----

emu_theses_bigrams_top <- emu_theses_bigrams %>%
  group_by(title) %>%
  count(bigram, sort = TRUE) %>%
  slice_max(n, n = 20) %>%        ### top_n is superseded; better use slice_max instead
  # filter(n > 50)                ### filter words with a minimum count
  ungroup() %>% 
  mutate(bigram = reorder(bigram, n))

@@

emu_theses_bigrams_top %>%
  filter(title == unique(title)[1:6]) |> 
  ggplot(aes(x = bigram, y = n, fill = title)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
  labs(y = "Count",
       x = "Unique words",
       title = "Count of unique words found in thesis",
       subtitle = "Stop words were removed from the list") +
  facet_wrap( ~ title, scales = "free", ncol = 2) +
  theme(legend.position = "none") 
```

```{r}
#| label: tf-idf

emu_theses_words

emu_theses_words_count <- emu_theses_words |> 
  count(title, word, sort = TRUE) |> 
  ungroup()

emu_theses_words_total <- emu_theses_words_count |> 
  group_by(title) |> 
  summarise(total = sum(n))

emu_theses_words_count <- left_join(emu_theses_words_count, emu_theses_words_total)

# emu_theses_words_count |> 
#   group_by(title) |> 
#   summarise(total = unique(tot

# Preview term frequency distributions in a subset of theses
ggplot(emu_theses_words_count |> 
         filter(title == unique(title)[1:6]),
       aes(n/total, fill = title)) +
  geom_histogram(show.legend = FALSE, binwidth = 0.0001) +
  xlim(NA, 0.0009) +
  facet_wrap(~ title, ncol = 3, scales = "free_y")

# Examine Zipf's law for the EMU theses
## Calculate rank and term frequency
freq_by_rank <- emu_theses_words_count |> 
  group_by(title) |> 
  mutate(rank = row_number(),
         `term frequency` = n/total)

## Visualise Zipf's law
freq_by_rank |> 
  ggplot(aes(rank, `term frequency`, color = title)) +
  geom_line(size = 1.1, alpha = 0.8, show.legend = FALSE) +
  scale_x_log10() +
  scale_y_log10()

rank_subset <- freq_by_rank |> 
  filter(rank < 500,
         rank > 10)

lm(log10(`term frequency`) ~ log10(rank), data =  rank_subset)

freq_by_rank |> 
  ggplot(aes(rank, `term frequency`, color = title)) +
  geom_line(size = 1.1, alpha = 0.8, show.legend = FALSE) +
  geom_abline(intercept = -1.44, slope = -0.7, color = "gray50", linetype = 2) +
  scale_x_log10() +
  scale_y_log10()

## Calculate tf-idf
emu_theses_words_count <- emu_theses_words_count |> 
  bind_tf_idf(word, title, n)
emu_theses_words_count

emu_theses_words_count |> 
  select(-total) |> 
  arrange(desc(tf_idf))

## Visualise the top n words with the highest tf-idf values for each of a subset of theses
emu_theses_words_count |> 
  filter(title == unique(title)[1:6]) |> 
  arrange(desc(tf_idf)) |> 
  mutate(word = factor(word, levels = rev(unique(word)))) |> 
  group_by(title) |> 
  top_n(15) |> 
  ungroup() |> 
  ggplot(aes(word, tf_idf, fill = title)) +
  geom_col(show.legend = FALSE) +
  labs(x = NULL, y = "tf-idf") +
  facet_wrap(~ title,  ncol = 3, scales = "free") +
  coord_flip()
```

The `tf-idf` statistic shows high values for the names of thesis locations that tend to be high-frequency thesis-specific words. As such, it shows words that are the most important to one document in a collection of documents [@silge2017].
